---
title: "Bus_Ridership_Data_Prep"
author: "Emmett"
date: "11/20/2020"
output: html_document
---

```{r}
library(dplyr)
library(httr)
library(jsonlite)
library(tidycensus)
library(purrr)
library(sf)
library(leaflet)
library(tidytransit)
library(ggplot2)
library(plotly)
library(tidyverse)
```

Part 1: Get DATA

Read in MBTA GTFS File
```{r}
##Read in MBTA GTFS##

gtfs <- read_gtfs("https://cdn.mbta.com/MBTA_GTFS.zip") 

View(gtfs$route_patterns)


gtfs_sf <- gtfs_as_sf(gtfs)
View(gtfs_sf$shapes)
View(gtfs_sf$routes)

View(gtfs_sf$trips)
lines <- get_route_geometry(gtfs_sf, service_ids = service_id_) %>% st_transform(4326)
st_write(lines, "bus_lines.geojson")
```



Select Inputs
```{r}
##Select Lines and Service You Want
route_id_select <- "47" ##input numeric later
service_id_select <- gtfs$calendar %>% pull(service_id) #switch for weekday/weekend service
period_select <- "AM_PEAK"
peak_select <- "PEAK"
day_type_select <- "weekday" ##note: only weekend service has "OFF_PEAK." Weekday service subdivided 
var_select <- "average_offs"
dir_select <- "0"
```


```{r}
## Build Map of Rail Lines

#rail_lines <- gtfs[["routes"]] %>% filter(route_desc %in% c("Rapid Transit","Commuter Rail")) %>%
#  left_join(lines, by = "route_id") %>%
#  mutate(route_color = paste0("#",route_color))

#rail_lines <- st_as_sf(rail_lines) %>% st_transform(4326) 

rail_pal <- colorFactor(rail_lines$route_color[0:21], rail_lines$route_id[0:21], ordered = TRUE, reverse = FALSE)

#t_stops <- filter_stops(gtfs, service_id_select, rail_lines$route_id) %>% 
#  select("stop_id","stop_lat","stop_lon", "stop_name","platform_name","wheelchair_boarding") %>% as.data.frame()

#t_stops_sf <- st_as_sf(t_stops, coords = c("stop_lon","stop_lat"), remove = FALSE, crs = 4326) 

st_write(t_stops_sf, "t_stops.geojson")
st_write(rail_lines, "rail_lines.geojson")

rail_lines <- st_read("rail_lines.geojson")
t_stops_sf <- st_read("t_stops.geojson")

## Add this to background of Bus Map

leaflet(t_stops_sf) %>% addProviderTiles(providers$CartoDB.DarkMatter) %>%
  setView(lng = -71.100003, lat = 42.340479, zoom = 13) %>%
  addCircleMarkers(lat = t_stops_sf$stop_lat, lng = t_stops_sf$stop_lon, radius = 3, fillColor = 'white', color = 'white', weight = 1.0) %>% 
  addPolylines(data = rail_lines, weight = 2, color = ~rail_pal(route_id), popup = paste(rail_lines$route_id))
  
```

```{r}
#stops <- gtfs$stops  #  vehicle_type %in% c("0","1","2","3"))

#zone_id == "RapidTransit" |

t_stops <- gtfs$stops %>% filter(grepl("CR",zone_id) | grepl("SL",zone_id) | grepl("RapidTransit",zone_id))

t_stops <- t_stops %>% mutate(color = 
                                if_else(grepl( "Orange", stop_desc), "#ED8B00", #orange line
                                  if_else(grepl("Red", stop_desc), "#DA291C", #red line
                                    if_else(grepl("Green", stop_desc), "#00843D",#green line
                                      if_else(grepl("Blue", stop_desc), "#003DA5", #blue line
                                        if_else(grepl("Commuter Rail", stop_desc), "#80276C", #commuter rail
                                            if_else(grepl("Silver Line", stop_desc), "#C9D3D8", "#FFFFFF") 
                                              )       
                                            )
                                          )
                                        )
                                      )
                                    ) %>%
  select(c("stop_id","stop_name","platform_name","stop_desc","stop_lat","stop_lon","wheelchair_boarding","color"))
                                   
t_stops_color <- st_as_sf(t_stops, crs = 4326, coords = c("stop_lon","stop_lat"), remove = FALSE)

plot(t_stops_color['color'])   

st_write(t_stops_color, "t_stops_color.geojson")
```


C) Route + Line Geometries
```{r}

##GET BUS STOPS DATA
stops <- gtfs$stops %>% 
  filter(vehicle_type == 3) %>%
  select("stop_id","stop_lat","stop_lon", "stop_name", "wheelchair_boarding") %>%
  mutate(stop_id = as.integer(stop_id)) %>% st_as_sf(remove = FALSE, crs = 4326, coords = c("stop_lon","stop_lat"))

st_write(stops, "bus_stops.geojson")

stops_tbl <- gtfs$stops %>% 
  filter(vehicle_type == 3) %>%
  select("stop_id","stop_lat","stop_lon", "stop_name", "wheelchair_boarding") %>%
  mutate(stop_id = as.integer(stop_id)) 

stops_sf <- st_as_sf(stops_tbl, coords = c("stop_lon", "stop_lat"), 4326) 

#%>%
#  st_transform(4326)

#Get human readable trip name
trips <- gtfs$trips %>% select(route_id, direction_id, trip_headsign) %>% distinct_at(c("route_id", "direction_id", "trip_headsign"))

write.csv(trips, "/Users/emmettmckinney/Documents/CodeAcademy/Bus_Ridership/data_for_viz/trips.csv")
```


A) Reliability Data (ROUTE LEVEL)
```{r}
## GET RELIABILITY DATA ## 

rel_route <- read.csv("/Users/emmettmckinney/Documents/CodeAcademy/Bus_Ridership/data_for_viz/MBTA_Reliability_Fall_2019.csv") %>%
  select(-c("gtfs_route_short_name","gtfs_route_long_name","metric_type","cancelled_numerator")) %>%
  filter(gtfs_route_id == route_id_select) %>% 
  mutate(on_time_rate = otp_numerator/otp_denominator) %>% #calculate on-time rate for date range
  mutate(service_date = substring(service_date, 0, 10)) %>%
  mutate(service_date = strptime(as.character(service_date), "%Y/%m/%d") %>% format("%Y-%m-%d"))

rel_overall <- read.csv("/Users/emmettmckinney/Documents/CodeAcademy/Bus_Ridership/data_for_viz/Daily_Bus_Reliability_Fall_2019.csv") %>%
  mutate(service_date = strptime(as.character(service_date), "%Y/%m/%d") %>% format("%Y-%m-%d") )##read in local file, calculated using code in "Overall_Reliability.Rmd" file

rel_chart <- rel_route %>% group_by(peak_offpeak_ind) %>%
  select(gtfs_route_id, on_time_rate, peak_offpeak_ind, service_date) %>%
  rename("rte_on_time_rate" = on_time_rate) %>% 
  left_join(rel_overall, by = c("peak_offpeak_ind","service_date")) %>% 
              rename("route" = rte_on_time_rate, "overall" = on_time_rate) %>%
              pivot_longer(rel_chart, 
              cols = c("route","overall"), 
              names_to = "type", 
              values_to = "on_time_rate") %>%
              mutate(type = as.factor(type))

```


B) RIDERSHIP DATA (STOP LEVEL)

```{r}

## GET RIDERSHIP DATA ## 
riders <- read.csv('/Users/emmettmckinney/Documents/CodeAcademy/Bus_Ridership/data_for_viz/MBTA_Bus_Ridership_by_Time_Period%2C_Season%2C_Route_Line%2C_and_Stop.csv') %>%
  filter(time_period_name == period_select,
         day_type_name == day_type_select,
         route_id == route_id_select,
         direction_id == dir_select
         )
```


Part 2: VISUALIZATIONS

```{r}
##LEAFLET MAP FOR BOARDINGS & ALIGHTINGS AT STOPS##

rider_viz <- riders %>% select(c("route_id","direction_id","day_type_name","time_period_name","stop_id",all_of(var_select),"num_trips","ObjectId")) %>% 
    rename_at(vars(starts_with('average')), funs(paste0('var'))) %>% #rename selected variable to standard column so leaflet can read it
  left_join(stops, by = "stop_id") %>%
  left_join(trips, by = c("route_id","direction_id"))

##Set Up Pallette

rider_pal <- colorBin("Greens", domain = rider_viz$var, bins = 5)

off_pal <- colorBin("Reds", domain = rider_viz$var, bins = 5)
on_pal <- colorBin("Greens", domain = rider_viz$var, bins = 5)
load_pal <- colorBin("Blues", domain = rider_viz$var, bins = 5)

#rider_pal <- if_else("average_ons" %in% as.character(var_select), on_pal, ## use a switch to build this in reactive object.
             # if_else("average_offs"%in% as.character(var_select), off_pal, load_pal))
                                                   
          
## Map Data 
var_txt <- if_else(var_select == "average_ons", 'Avg. Boardings',
                   if_else(var_select == "average_offs", "Avg. Alightings", "Avg. Load"))

leaflet(lines %>% filter(route_id == route_id_select)) %>%
            setView(lng = -71.100003, lat = 42.340479, zoom = 13) %>%
            addProviderTiles(providers$CartoDB.DarkMatter,
                             options = providerTileOptions(opacity = 0.8))  %>%
            addPolylines(color = "white", weight = 1) %>%
            addCircleMarkers(lng = rider_viz$stop_lon, lat = rider_viz$stop_lat, radius = rider_viz$var, fillColor = ~rider_pal(rider_viz$var), 
                             color = ~rider_pal(rider_viz$var), 
                             fillOpacity = 0.5, 
                             popup = paste0("Stop: ", rider_viz$stop_name, "<br>", var_txt, " ", rider_viz$var, "<br>", "Destination: ", rider_viz$trip_headsign), 
                             options = popupOptions(autoPan = TRUE, riseOnHover = TRUE))

```

```{r}
##LEAFLET MAP FOR RELIABILITY ON LINE##

rel_viz <- rel_route %>% left_join(lines, by = c("gtfs_route_id" = "route_id"))
rel_viz <- rel_viz %>% st_as_sf

rel_pal <- colorBin("RdYlGn", domain = reel_viz$on_time_rate, bins = 5) 

rel_summary <- rel_viz %>% group_by(gtfs_route_id) %>% summarize(
  on_time_rate = mean(on_time_rate))

leaflet(rel_summary) %>%
            setView(lng = -71.100003, lat = 42.340479, zoom = 13) %>%
            addProviderTiles(providers$CartoDB.DarkMatter,
                             options = providerTileOptions(opacity = 0.8))  %>%
            addPolylines(color = ~rel_pal(on_time_rate), weight = 2)
      
```

```{r}
## VIOLIN PLOT OF ROUTE RELIABILITY Relative to others ##

ggplot(rel_chart, aes(fill = type, color = type, x = type, y = on_time_rate)) + 
  geom_violin(position="dodge", alpha=0.5, outlier.colour="transparent",
              draw_quantiles = c(0.25, 0.5, 0.75)) +
  geom_jitter(height = 0, width = 0.05, color = "black") +
  
  ylab("On Time %") +
  theme(legend.title = element_blank()) +
  xlab(element_blank())
#v

  
```

```{r}
## ROUTE PROFILE 

rider_chart <- riders %>% select(c("ObjectId", "route_id","stop_id","stop_sequence","stop_name","average_ons","average_offs")) %>% 
 pivot_longer(cols = c("average_ons","average_offs"), names_to = "type", values_to = "count")

rider_load <- riders %>% select(c("ObjectId", "route_id","stop_id","stop_sequence","stop_name","average_load")) %>% 
 pivot_longer(cols = "average_load", names_to = "type", values_to = "count")

p <- ggplot(rider_chart, (aes(x = stop_sequence, y=count, text = stop_name, fill=type))) +
        geom_bar(position = "stack", stat="identity") +
        geom_step(data = rider_load, color = "#70dcf0", size = 0.8, group = 1)  +
        scale_fill_manual(values = c("#70dcf0","#f768a1","#66c2a4"), labels = c("Pass. Load", "Alightings","Boardings")) +
  theme(legend.title = element_blank(), panel.background = element_blank()) +
    xlab("Stop Sequence") +
    ylab("Count")

p
```


```{r}
p

fancy <- ggplotly(p, tooltip = c("count","stop_name"))

fancy

```




Ideas:: 

DONE __ Passenger Crowding: Identify route segments based on average load at stops where the passenger load is highest

DONE ---- Overview: download reliability data for all lines, summarize over whole time period, group by peak / off-peak, direction, turn rail lines on & off.

Default leaflet plot has all bus routes at very low opacity, the 1 Route highlighted.

Botton left panel: Ridership plot showing the route profile over the course of the route (stacked bar chart like BlueBike Viz)

Botton right panel: Reliability plot (violin/box plot showing reliabilitiy, day-to-day, vs. average for all routes in that period. Will need to download local summary of all routes for Fall 2019 to do this.)

Right side panel:

Controls / Instructions:
 - select routes
 - select peak / off-peak

statistics: 
  - "You've selected Route #86, 
  -  Direction
  -  Headways,
  -  bus type, 
  -  demographics of riders (CTPS Ridership Survey?), 
  -  reliability ranking
  -  designation as a key route.
  
  - You've selected stop: (on click, other routes that stop there also highlight)
  - Name: Washington St @ Sq.
  - Routes that stop here:
  


Part 2: Summarizing Data

```{r}
##SPATIAL ANALYSIS OF BUS ROUTES## — figure out demographics of areas around line
```

```{r}
##SPATIAL ANALYSIS OF BUS STOPS - figure out demographics of areas around bus stops
```

```{r}
##SUMMARIZE STOP DATA
```

```{r}
#SUMMARIZE RELIABILITY DATA
```

